Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2018-05-26 paper.tex
Add Gail's notes from May 23 email

Pinching parameters
(calculated by eye from the \emph{spectral} \emph{scat} method)
for the species
$\{\nu_e,\bar{\nu}_e,\nu_x\}$ are $\alpha\approx\{3.6,5.1,2.3\}$.
Energy moments of pinched spectra like those in Eqn.~\ref{eqn:avg_eps_sc}
have a simple analytic form so that Eqn.~\ref{eqn:avg_eps_sc} becomes
\begin{align}
  \label{eqn:pinched_avg_eps_sc_1}
  \langle\tilde{\varepsilon}\rangle^{\rm sc}
  &\approx \frac{\alpha+3}{\alpha+1} \langle\varepsilon\rangle, \\
  %\label{eqn:pinched_avg_eps_sc_2}
  %&= \{1.4,1.3,1.6\} \langle\varepsilon\rangle, \\
  \label{eqn:pinched_avg_eps_sc_3}
  &\approx \{16,18,28\}\,~{\rm MeV},
\end{align}
again for $\{\nu_e,\bar{\nu}_e,\nu_x\}$ respectively.
These analytic predictions agree with the average energies of the scattering
envelope to approximately 10\%
% From by-eye examinations of <eps>(cosA) plot using data from
% MCDATA=$DIR/ResultsSherwood-02
% RTDATA_GR=$DIR/ResultsBrett-B-03/Test-scat-01/
% RTDATA_SP=$DIR/ResultsBrett-A-03/Test-scat-01/
%          nue     nua     nux
of all of the treatments including elastic scattering presented
in Tab.~\ref{tab:collapse_rt_vs_mc}, except for
$\bar{\nu}_e$ in the \emph{gray} \emph{scat} treatment,
which deviates from our prediction by 30\%.
This agreement is excellent despite the drastic
simplifications used in our toy model.

@@

Fig.~\ref{fig:nsns_theta_distrib_avg_eps} shows the distribution of
average energies of radial fluxes over observer positions.
With and without scattering, the average energies are highest
for observers near the polar axis, since polar observers get a
direct view of the hot hypermassive neutron star.
(This trend is unexpectedly reversed for $\nu_x$,
and may be due to asymetries in the disk, as discussed above.)
Scattering decreases average energies across all observer positions,
as it did in the post-bounce configuration, or any configuration of
a hot interior surrounded by a scattering envelope.
The strength of the effect of scattering on the average energies of
the different species is seen to follow the ranking
$\nu_x>\bar{\nu}_e>\nu_e$ due to two factors:
the average energies of the spectra obey the same ranking,
and the thicknesses of the different species' scattering envelopes
also obey the same ranking.
As we showed in the toy models with a homogeneous background scattering field,
the scattering contribution $f_{\rm SE}$ is
proportional to energy $\varepsilon^2$ and path length $s$
(Eqn.~\ref{eqn:fse_homogeneous_thin_limit}).
\begin{figure}
  %\includegraphics[width=\columnwidth]{theta_distrib-250km-avg_eps}
  \resizebox{\columnwidth}{!}{\input{theta_distrib-250km-avg_eps.tex}}
  % see notes from theta_distrib-250km-Kr
  % Script: $DIR/GnuScripts/plot-theta_distributions-avg_eps_from_fluxes-fixed_r.gnu
  \caption{Same as Fig.~\ref{fig:nsns_theta_distrib_Kr} but
    average energies of radial fluxes
    $\langle\tilde{\varepsilon}\rangle(\theta)=\tilde{H}^r(\theta)/\tilde{K}^r(\theta)$,
    as defined in Eqns.~\ref{eqn:Kr_minkowski} and \ref{eqn:Hr_minkowski}.
  }
  \label{fig:nsns_theta_distrib_avg_eps}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{fig-survival_prob-scat_ih}
  % Flavor evolution by Yonglin, see emails 5.9.18, 5.16.18
  % Data: $DIR/300-1499kmACC12Step2NE30IH051418m12sc.dat2p
  % Script: $DIR/plot.gp
  % w/ DIR=/Users/mbdeaton/Documents/ResearchGraveyard/NuclearAstrophysics/DataAndVis/20170412-self_interaction_potentials_M12_M12-LS220-M1/Run03/RunBA/RunBA-025deg-scat/flavor_evolution
  \caption{Same as Fig.~\ref{fig:survival-noscat},
    but for the \emph{scat} case.
    In this case the neutrinos undergo a standard MNR transition
    beginning around $r\sim400$~km,
    with the electron neutrinos converting to mu or tau neutrinos almost
    completely,
    and the electron antineutrinos oscillating back to their original flavor
    after partially converting to mu and tau.
  }
  \label{fig:survival-scat}
\end{figure}

@@

We have also employed the ray tracing code to examine neutrino flavor
oscillations along one toy trajectory exiting the neutrino-dense environment
of the neutron star post-merger configuration (Sec.~\ref{ssec:V_nunu}).
The trajectory starts from 300~km, and moves out radially at
$25^\circ$ from the polar axis.
Along that trajectory, elastic scattering has the effect of increasing
the ratio of $\bar{\nu}_e$ relative to $\nu_e$.
This creates a negative self-interaction potential
which introduces a complete standard matter neutrino resonance
(see Fig.~\ref{fig:survival-scat}).
At about 400~km from the merger core,
both electron neutrinos and electron antineutrinos begin to transform.
At about 1200~km
electron neutrinos have almost completely converted to mu or tau neutrinos,
while the electron antineutrinos have returned back to their original flavor.
In an otherwise identical calculation, ignoring elastic scattering
causes the flavor transformation to be very different (see Fig.~\ref{fig:survival-noscat}).

@@

\subsubsection*{Acknowledgements}
This research is supported in part by the Joint Institute for Nuclear
Astrophysics---Center for the Evolution of the Elements (JINA-CEE)
and the National Science Fondation under grant numbers X. \footnote{TODO (all): grant nos?}
We thank Sherwood Richers for providing his neutrino transport code
\lstinline{Sedonu} for comparisons,
Fran\c{c}ois H\'{e}bert for his timely help in finding and
correcting bugs in the time-interpolation algorithm,
and Luke Roberts for helpful discussions.

>>>>

Pinching parameters
(calculated by eye from the \emph{spectral} \emph{scat} method)
for the species
$\{\nu_e,\bar{\nu}_e,\nu_x\}$ are $\alpha\approx\{3.6,5.1,2.3\}$.
Energy moments of pinched spectra like those in Eqn.~\ref{eqn:avg_eps_sc}
have a simple analytic form so that Eqn.~\ref{eqn:avg_eps_sc} becomes
\begin{align}
  \label{eqn:pinched_avg_eps_sc_1}
  \langle\tilde{\varepsilon}\rangle^{\rm sc}
  &\approx \frac{\alpha+3}{\alpha+1} \langle\varepsilon\rangle, \\
  %\label{eqn:pinched_avg_eps_sc_2}
  %&= \{1.4,1.3,1.6\} \langle\varepsilon\rangle, \\
  \label{eqn:pinched_avg_eps_sc_3}
  &\approx \{16,18,28\}\,~{\rm MeV},
\end{align}
again for $\{\nu_e,\bar{\nu}_e,\nu_x\}$ respectively.
These analytic predictions agree with the average energies of the scattering
envelope to approximately 10\%
% From by-eye examinations of <eps>(cosA) plot using data from
% MCDATA=$DIR/ResultsSherwood-02
% RTDATA_GR=$DIR/ResultsBrett-B-03/Test-scat-01/
% RTDATA_SP=$DIR/ResultsBrett-A-03/Test-scat-01/
%          nue     nua     nux
of all of the treatments including elastic scattering presented
in Tab.~\ref{tab:collapse_rt_vs_mc}, except for
$\bar{\nu}_e$ in the \emph{gray} \emph{scat} treatment,
which deviates from our prediction by 30\%.
This agreement is excellent despite the drastic
simplifications used in our model.

@@

Fig.~\ref{fig:nsns_theta_distrib_avg_eps} shows the distribution of
average energies of radial fluxes over observer positions.
With and without scattering, the average energies are highest
for observers near the polar axis, since polar observers get a
direct view of the hot hypermassive neutron star.
(This trend is unexpectedly reversed for $\nu_x$,
and may be due to asymetries in the disk, as discussed above.)
Scattering decreases average energies across all observer positions,
as it did in the post-bounce configuration, or any configuration of
a hot interior surrounded by a scattering envelope.
The strength of the effect of scattering on the average energies of
the different species is seen to follow the ranking
$\nu_x>\bar{\nu}_e>\nu_e$ due to two factors:
the average energies of the spectra obey the same ranking,
and the thicknesses of the different species' scattering envelopes
also obey the same ranking.
As we showed in the models with a homogeneous background scattering field,
the scattering contribution $f_{\rm SE}$ is
proportional to energy $\varepsilon^2$ and path length $s$
(Eqn.~\ref{eqn:fse_homogeneous_thin_limit}).
\begin{figure}
  %\includegraphics[width=\columnwidth]{theta_distrib-250km-avg_eps}
  \resizebox{\columnwidth}{!}{\input{theta_distrib-250km-avg_eps.tex}}
  % see notes from theta_distrib-250km-Kr
  % Script: $DIR/GnuScripts/plot-theta_distributions-avg_eps_from_fluxes-fixed_r.gnu
  \caption{Same as Fig.~\ref{fig:nsns_theta_distrib_Kr} but
    average energies of radial fluxes
    $\langle\tilde{\varepsilon}\rangle(\theta)=\tilde{H}^r(\theta)/\tilde{K}^r(\theta)$,
    as defined in Eqns.~\ref{eqn:Kr_minkowski} and \ref{eqn:Hr_minkowski}.
  }
  \label{fig:nsns_theta_distrib_avg_eps}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{fig-survival_prob-scat_ih}
  % Flavor evolution by Yonglin, see emails 5.9.18, 5.16.18
  % Data: $DIR/300-1499kmACC12Step2NE30IH051418m12sc.dat2p
  % Script: $DIR/plot.gp
  % w/ DIR=/Users/mbdeaton/Documents/ResearchGraveyard/NuclearAstrophysics/DataAndVis/20170412-self_interaction_potentials_M12_M12-LS220-M1/Run03/RunBA/RunBA-025deg-scat/flavor_evolution
  \caption{Same as Fig.~\ref{fig:survival-noscat},
    but for the \emph{scat} case.
    In this case the neutrinos undergo a standard MNR transition
    beginning around $r\sim400$~km,
    with the electron neutrinos converting to mu or tau neutrinos almost
    completely,
    and the electron antineutrinos oscillating back to their original flavor
    after partially converting to mu and tau antineutrinos.
  }
  \label{fig:survival-scat}
\end{figure}

@@

We have also employed the ray tracing code to examine neutrino flavor
oscillations along one sample trajectory exiting the neutrino-dense environment
of the neutron star post-merger configuration (Sec.~\ref{ssec:V_nunu}).
The trajectory starts from 300~km, and moves out radially at
$25^\circ$ from the polar axis.
Along that trajectory, elastic scattering has the effect of increasing
the ratio of $\bar{\nu}_e$ relative to $\nu_e$.
This creates a negative self-interaction potential
which introduces a complete standard matter neutrino resonance transition
(see Fig.~\ref{fig:survival-scat}).
At about 400~km from the merger core,
both electron neutrinos and electron antineutrinos begin to transform.
At about 1200~km
electron neutrinos have almost completely converted to mu or tau neutrinos,
while the electron antineutrinos have returned back to their original flavor.
In an otherwise identical calculation, ignoring elastic scattering
causes the flavor transformation to be very different (see Fig.~\ref{fig:survival-noscat}).

@@

\subsubsection*{Acknowledgements}
This research is supported in part by the Joint Institute for Nuclear
Astrophysics---Center for the Evolution of the Elements (JINA-CEE)
and the National Science Foundation under grant numbers X,
and the U.S.\ Department of Energy, Office of Science, Office of Nuclear Physics,
under Award Number DE-FG02-02ER41216.
\footnote{TODO (all): grant nos?}
We thank Sherwood Richers for providing his neutrino transport code
\lstinline{Sedonu} for comparisons,
Fran\c{c}ois H\'{e}bert for his timely help in finding and
correcting bugs in the time-interpolation algorithm,
and Luke Roberts for helpful discussions.
